---
title: "Code PS IPTW categorical variable"
author: "Jeanne LECLERE"
date: today 
format: html
editor: visual
warning: FALSE
message: FALSE
---

### Packages

```{r}
library("dplyr")
library("tidyverse")
library("cobalt")
library('ggplot2')
library('gtsummary')
```

## Data formatting and creation of a categorical variable

We calculate the average SEMSA score and then construct our categories by choosing the corresponding percentiles: 3 categories, 100/3 for each percentile

```{r}
#Opening file
data <- read.csv("donnees_synthetiques.csv", header = TRUE, sep=",") 

#Search for values corresponding to percentiles to create categories
data |>
  tbl_summary(
    include = c(SEMSA_TOT),
    statistic = 
      all_continuous() ~ "Percentile : {p33} {p66} {p99} ")
p33 <- quantile(data$SEMSA_TOT, probs = 0.33, na.rm = TRUE)
p66 <- quantile(data$SEMSA_TOT, probs = 0.66, na.rm = TRUE)

#Creation of the HLQ categorical score variable: 2,3,4 becomes bad, average, good
data <- data %>%
  mutate(
    SEMSA_cat = case_when(
  SEMSA_TOT <= p33  ~  "bad",
  SEMSA_TOT > p33  & SEMSA_TOT <= p66 ~ "average",
  SEMSA_TOT > p66  ~ "good")
  )

#Transform the characters into categories with 3 levels
data$SEMSA_cat <- factor(data$SEMSA_cat, levels = c("bad", "average", "good"))
```

### Calculation of the multinomial regression of the SEMSA score function of covariates to calculate the propensity score

The score is calculated using multinomial logistic regression (SEMSA variable is transformed into a categorical variable, resulting in a score of 2 -\> bad, 3 -\> average, 4 -\> good). It corresponds to the conditional probability (y) of exposure, weighted by the presence of covariates (x) (defined by DAG).

Here, the chosen covariates are age, gender, and BMI, for example.

```{r}
#Calcul PS par regression logistique multinomial
ps_model <- nnet::multinom(SEMSA_cat ~ AGE + GENDER + BODY.MASS.INDEX + APNEA.HYPOPNEA.INDEX,
                data = data)
coefs <- coef(ps_model)
print(coefs)
summary(ps_model)

#Visualization
ps_model |> 
  ggstats::ggcoef_dodged(exponentiate = TRUE)
```

Threshold between the average/poor and good/poor categories; poor is the reference threshold.

We observe that increasing age increases the odds ratio (OR): exp(0.00017) = 1.00 and exp(0.0274) = 1.03. The older the age, the higher the probability of having a better HLQ score (3% difference between bad and good).

For gender, the OR = exp(-0.52) = 0.59 and exp(-0.67) = 0.51 (1 being coded for males). Men have a 41/49% lower probability of being average/good versus bad.

For BMI, the OR exp(-0.0006) = 0.99 and exp(0.02) = 1.02, the higher the weight, the higher the probability of having a better HLQ score (+2% between bad vs. good).

### Calculating the propensity score manually and verifying the distribution

We obtain the following equation:

log(P(average)/P(reference = bad)) = β0 + β1*AGE +* β2\*GENDER + β3\*BMI

where k = 2 thresholds (Threshold 1 = we compare mean vs. bad and Threshold 2 = we compare good vs. bad) from which we calculate the propensity score of each individual added to the database: (Knowing that the sum of the probabilities = 1)

PS_bad = 1 / (1+e^naverage^+e^ngood^) où naverage = β0 + β1\*Age + β2\*Gender + β3\*BMI etc.

PS_average= e^naverage^ / (1+e^naverage^+e^ngood^ )

PS_good = e^ngood^/ (1+e^naverage^+e^ngood^ )

We calculate the probability of being in each category for each patient in order to estimate the entire distribution according to the chosen covariates.

```{r}
#Calculating probabilities for each category
proba_average <- coefs["average","(Intercept)"] + coefs["average","AGE"] * data$AGE + 
  coefs["average","GENDER"] * data$GENDER + coefs["average","BODY.MASS.INDEX"] * data$BODY.MASS.INDEX + coefs["average","APNEA.HYPOPNEA.INDEX"] * data$APNEA.HYPOPNEA.INDEX

proba_good <- coefs["good","(Intercept)"] + coefs["good","AGE"] * data$AGE + 
  coefs["good","GENDER"] * data$GENDER + coefs["good","BODY.MASS.INDEX"] * data$BODY.MASS.INDEX + coefs["good","APNEA.HYPOPNEA.INDEX"] * data$APNEA.HYPOPNEA.INDEX

exp_average <- exp(proba_average)
exp_good   <- exp(proba_good)

denom <- 1 + exp_average + exp_good

data$ps_bad <- 1 / denom
data$ps_average <- exp_average / denom
data$ps_good <- exp_good / denom

#Sum of scores = 1 for each individual
summary(data$ps_bad)
summary(data$ps_average)
summary(data$ps_good)

#We isolate each PS for each calculated individual
data <- data %>%
  mutate(
    ps_ind = case_when(
  SEMSA_cat == "bad" ~ data$ps_bad,
  SEMSA_cat == "average" ~ data$ps_average,
  SEMSA_cat == "good" ~ data$ps_good)
  )
head(table(data$ps_ind))

#Visualization
data %>%
  ggplot(aes(x = ps_ind, fill = SEMSA_cat)) +
  geom_density(alpha = 0.5) +  scale_x_log10() +
  scale_fill_manual(values= c("bad"="purple","average"="yellow","good"="blue"))
```

### Calculating the propensity score with PREDICT and verifying the distribution

We calculate the probability of being in each category for each patient in order to estimate the entire distribution according to the chosen covariates.

```{r}
#Calculating the propensity score using predict = probability of being in each category for all individuals
table_ps <- data.frame(predict(ps_model, data, type = "probs")) 
head(table_ps)

#Add the PS to the database
data <- cbind(data, table_ps)

#Verification of PS distribution
summary(table_ps)

#Visualization of PS according to the categories
#Pivot columns to visualize the data
data_graph <- data  %>% 
  select(bad, average, good) %>%
  pivot_longer(cols = everything(), names_to = "Categories", values_to = "Score")

#Visualization
ggplot(data_graph, aes(x = Score, fill = Categories)) +
  geom_density(alpha = 0.5) +  scale_x_log10() +
  scale_fill_manual(values= c("bad"="purple","average"="yellow","good"="blue"))
```

### Calculating the IPTW score

The IPTW score is calculated from the propensity score and corresponds to the inverse of the propensity score for each individual.

For a categorical variable, the weight depends on the probability of the observed category; it corresponds to w_i = 1 / (P(T_i = t_i \| X_i) =\> 1/PS calculated previously.

```{r}
#Calculation of frequency to calculate stabilized weights
data <- data %>%
  group_by(SEMSA_cat) %>%
  mutate(proba_exposure_cat = n() / nrow(data)) %>%
  ungroup()

#Weight calculation = inverse of propensity score with individual weight. DO NOT ROUND.
data <- data %>%
  mutate(
    iptw = case_when(
  SEMSA_cat == "bad" ~ proba_exposure_cat/data$ps_ind,
  SEMSA_cat == "average" ~ proba_exposure_cat/data$ps_ind,
  SEMSA_cat == "good" ~ proba_exposure_cat/data$ps_ind))
head(table(data$iptw))

#Visualization of the balance of covariances after weighting
graph <- bal.tab(SEMSA_cat ~ AGE + GENDER + BODY.MASS.INDEX + APNEA.HYPOPNEA.INDEX,
                 weights = data$iptw,
                 threshold = 0.1,
                 data = data,
                 un = TRUE)
print(graph)
love.plot(graph)
```

### Verification of the distribution of IPTW scores in the population

```{r}
#Verification of weight distribution
summary(data$iptw)

#Visualization
data %>%
  ggplot(aes(x = iptw, fill = SEMSA_cat)) +
  geom_density(alpha = 0.5) +  scale_x_log10() +
  scale_fill_manual(values= c("bad"="purple","average"="yellow","good"="blue"))
```

### Estimation of estimand

Here, the estimand is the ATE due to the use of the IPTW. It answers the research question: what is the adjusted effect of the covariates (age, gender, BMI) on the probability of treatment adherence function of the SEMSA score.

We use binomial regression to estimate treatment adherence (y = final outcome) function of the HLQ score, based on the previously calculated weights.

```{r}
#Regression with the new weights using binary regression to determine if membership = yes/no
fit <- glm(obs_J365_1M_PPC ~ SEMSA_cat, #effect of the HLQ score on binary variable adherence
           data = data,
           family = gaussian(), #corresponds to Y type
           weights = iptw)

summary(fit)
```

The p-value is significative \<0.05 for both categories compared to the reference : exemple for interpretation = atients with SEMSA_cat = average got exp(0.74) more odds to be adherent compared to SEMSA_cat = bad
