---
title: "Code PS IPTW continuous variable"
author: "Jeanne LECLERE"
date: today 
format: html
editor: visual
warning: FALSE
message: FALSE
---

### Packages used

```{r}
library("dplyr")
library("tidyverse")
library("gtsummary")
library("cobalt")
library('ggplot2')
```

## Data formatting and creation of a continuous variable

```{r}
#Opening file
data <- read.csv("donnees_synthetiques.csv", header = TRUE, sep=",")

table_sum <- 
  tbl_summary(
    data,
    include = c("BODY.MASS.INDEX", "GENDER", "AGE", "SEMSA_TOT", "APNEA.HYPOPNEA.INDEX"),
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "always",
    digits = all_continuous() ~ 2
  ) |>
  modify_header(
    label = "**Variable**"
  )
print(table_sum)

#Visualization of SEMSA score
data_1 <- data
summary(data_1$SEMSA_TOT)
data_1 %>%
  ggplot(aes(x = SEMSA_TOT)) + 
  geom_density(fill = "lavender",alpha = 0.5) +  
  labs(title = "Distribution of SEMSA score") +
  theme_minimal() 
```

### Calculation of the regression of the SEMSA score as a function of covariates to calculate the propensity score

The score is calculated using linear regression. It corresponds to the conditional probability (y) of exposure, weighted by the presence of covariates (x) (defined by DAG).

Here, the chosen covariates are age, gender, and BMI, as examples.

```{r}
#Estimating continuous exposure with linear regression
ps_model <- lm(SEMSA_TOT ~ GENDER + AGE + BODY.MASS.INDEX + APNEA.HYPOPNEA.INDEX,
                data = data_1)
coefs <- coef(ps_model)
print(coefs)
summary(ps_model)
```

Here we have a linear regression, so we don't interpret the results using odds ratios (ORs) but only the p-values.

We observe that increasing age increases the probability of having a better HLQ score (p-value \< 0.05 (0.0038)).

Regarding gender, men have, on average, a lower SEMSAscore than women (p-value of 0.09 \> 0.05).

For BMI, the effect is not significant (p-value \> 0.05).

### Verification of the normality of distribution

The propensity score becomes a continuous density function characterized by a mean and a variance. It has three properties: the total area is 1, so it should be 0; the probability corresponds to the area, not the height.

It is necessary to calculate the regression residuals, which correspond to the difference between the observed and estimated values, and to verify their normality using a Shapiro-Mitsubishi test (check the conditions for use).

```{r}
#Visaul checking of normal distribution 
qqnorm(residuals(ps_model))
qqline(residuals(ps_model))
```

### Estimating the conditional exposure

```{r}
#Estimating the parameters (mean, variance) of normaal distribution 
score <- predict(ps_model) 
sigma <-sd(residuals(ps_model))

#Estimation of conditional density of GPS
data_1$gps <-dnorm(data_1$SEMSA_TOT, mean=score, sd=sigma) 
summary(data_1$gps) 

#Visualization
data_1 %>%
  ggplot(aes(x = gps)) + 
  geom_density(fill = "lavender",alpha = 0.5) +  
  labs(title = "Distribution of GPS") +
  theme_minimal() 
```

### Reprensentation of a patient's gps (density) with his SEMSA score observed

```{r}
#Patient i
i <- 5

#Creation 
#Predicted value
mean <- score[i] 
sigma

#Observed value
T_i <- data_1$SEMSA_TOT[i] 

#Grid of T values with 100 points on the curve
t_vals <- seq(min(data_1$SEMSA_TOT),
              max(data_1$SEMSA_TOT),
              length.out = 100) 

#Conditional density
dens <- dnorm(t_vals, mean = mean, sd = sigma)
df_plot <- data.frame(t = t_vals, dens = dens)

#GPS : Conditionnal density of the value
gps_i <- dnorm(T_i, mean = mean, sd = sigma)

ggplot(df_plot, aes(x = t, y = dens)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_vline(xintercept = T_i, color = "black") +
  annotate("point",x = T_i,y = gps_i, color = "red", size = 3) +
  theme_minimal() +
  labs(title = paste("Conditionnal density of patient", i),
       x = "SEMSA_tot",
       y = "Density")
```

### Caculation of IPTW weight

The IPTW score is calculated from the propensity score and is the inverse of the propensity score for each individual.

For a continuous variable, the weight depends on the observed conditional density function and is calculated as follows:

w_i = 1 / f(T_i \| X_i) =\> 1/GPS calculated previously.

```{r}
#Calculation of marginal density 
d_marg <- dnorm(data_1$SEMSA_TOT, mean = mean(data_1$SEMSA_TOT), sd = sd(data_1$SEMSA_TOT))

#Calculation stabilized weights
data_1$iptw <- d_marg/data_1$gps

#Visualization
data_1 %>%
  ggplot(aes(x = iptw)) + scale_x_log10() +
  geom_density(fill = "lavender",alpha = 0.5) +  
  labs(title = "Distribution of weights (log scale)") +
  theme_minimal() 
```

#### Trimming extreme weights

```{r}
#Trimming extreme weights through 5e and 95e quantiles
data_1$iptw_trim <- pmin(pmax(data_1$iptw, quantile(data_1$iptw, 0.05)), quantile(data_1$iptw, 0.95))

#Visualisation
data_1 %>%
  ggplot(aes(x = iptw_trim)) + scale_x_log10() +
  geom_density(fill = "lavender",alpha = 0.5) +  
  labs(title = "Distribution of trimmed weights (log scale)") +
  theme_minimal() 

#nombre de patient
data_1 <- data_1 %>%
  mutate(
    nb_patient_trim = case_when(
  iptw == iptw_trim ~ 0,
  iptw != iptw_trim ~ 1)
  )
sum(data_1$nb_patient_trim)
```

```         
```

### Verification of GPS distribution through exposition strata

```{r}
#creating quintiles of exposition 
data_1$SEMSA_q <- ntile(data_1$SEMSA_TOT, 5)

#Visualisation
data_1 %>%
  ggplot(aes(x = gps, fill = as.factor(SEMSA_q))) +
  geom_density(binwidth = 2,
                 position = "identity",
                 alpha = 0.5) +
  scale_fill_manual(values = c("purple", "yellow", "orange", "blue", "green")) +
  labs(x = "Quintiles of exposition", y = "Density of Gneralized Propensity Score", fill = "Quintiles of GPS") +  theme_classic()
```

#### Verification of the covariate balance after using the weights

```{r}
#Visualization
graph <- bal.tab(SEMSA_TOT ~ GENDER + BODY.MASS.INDEX + AGE + APNEA.HYPOPNEA.INDEX,
                 weights = data_1$iptw_trim,
                 data = data_1,
                 threshold = 0.1,
                 abs = TRUE,
                 un = TRUE)
print(graph)
love.plot(graph)
#to save :ggsave("distribution_GPS.png",plot = last_plot(),width = 6,height = 4,dpi = 300)
```

### Estimation of the estimand

Here, the estimand is the ATE due to the use of the IPTW. It answers the research question: what is the adjusted effect of the covariates (age, gender, BMI) on the treatment adherence function of the SEMSA score.

We use binomial regression to estimate treatment adherence (y = final outcome) function of the SEMSA score, based on the previously calculated weights.

#### Linear outcome

```{r}
#Linear regression with new weights to calculate estimand
fit <- glm(obs_J90_1M_PPC ~ SEMSA_TOT,
           data = data_1,
           family = gaussian(),
           weights = iptw_trim) 

summary(fit)
```

A 1-unit increase in the SEMSA score is associated with a 68-unit increase in PPC.

The p-value is \< 0.05, which shows that the model is statistically significant in predicting treatment adherence based on the total SEMSA score.

#### Binomial outcome

```{r}
#Logistic regression with new weights to calculate estimand
fit <- glm(obs_J365_periode ~ SEMSA_TOT,
           data = data_1,
           family = binomial(), 
           weights = iptw) 

summary(fit)
```

A 1-unit increase in the SEMSA score is associated with an e\^1.16 increase in the likelihood of experiencing event 1 in obs_J365_periode.

The p-value is \< 0.05, indicating that the model is statistically significant in predicting treatment adherence based on the total SEMSA score.

```{r}
ggplot(data_1, aes(x = SEMSA_TOT, y = obs_J90_1M_PPC)) +
  geom_point(alpha = 0.4) +
  geom_smooth(aes(weight = iptw), method = "loess", se = TRUE, color = "red") +
  #True = Ic 95% de la courbe rouge
  theme_minimal() +
  labs(x = "SEMSA score",
       y = "Prediction of adherence")
```
