---
title: "Code PS IPTW binary variable"
author: "Jeanne LECLERE"
date: today 
format: html
editor: visual
warning: FALSE
message: FALSE
---

### Packages

```{r}
library("dplyr")
library("tidyverse")
library("ggplot2")
```

## Data formatting and creation of a binary variable

```{r}
#Opening file
data <- read.csv("donnees_synthetiques.csv", header = TRUE, sep=",") 

#Visualization of distriubtion exciting variable
str(data$SEMSA)
hist(data$SEMSA_seuil)
```

### Calculation of logistic regression of the HLQ score function of covariates for the propensity score

The score is calculated using binary logistic regression because the HLQ variable is transformed into a binary variable. It corresponds to the conditional probability (y) of exposure, weighted by the presence of the covariates (x) (defined by DAG).

Here, the chosen covariates are age, gender, and BMI, for example.

```{r}
#Calculation of the propensity score by logistic regression 
ps_model <- glm(SEMSA_seuil ~ AGE + GENDER + BODY.MASS.INDEX + APNEA.HYPOPNEA.INDEX,
                data = data,
                family = binomial)
coefs <- coef(ps_model)
summary(ps_model)
```

We observe that increasing age increases the odds ratio (OR): exp(0.024) = 1.024. The older the age, the higher the probability of having a better HLQ score.

For gender, the OR = exp(-0.318) = 0.7 (1 being coded for males). Men have a lower probability of having a low HLQ score.

For BMI, the OR exp(0.011) = 1.011. The probability of having a better HLQ score increases with weight.

We obtain the logit(BMI) equation = β0 + β1*AGE + β2*GENDER + β3\*BMI (β0 = intercept), from which we calculate the propensity score of each individual added to the database :

PS = e\^logit / (1+e\^(logit))

### Calculating the propensity score

```{r}
#Calculation of the propensity score = Probability per each individual 
logit <- coefs["(Intercept)"] + coefs["AGE"] * data$AGE + coefs["GENDER"] * data$GENDER + coefs["BODY.MASS.INDEX"] * data$BODY.MASS.INDEX + coefs["APNEA.HYPOPNEA.INDEX"] * data$APNEA.HYPOPNEA.INDEX
#Propensity score
PS <- exp(logit) / (1 + exp(logit))
head(PS)

#Adding the score to the data
data$propensity_score <- PS
```

### Verification of the distribution of propensity scores in the population

```{r}
#Verification of the distribution of propensity scores
summary(data$propensity_score)

#Visualisation 
data %>%
  ggplot(aes(x = propensity_score, fill = as.factor(SEMSA_seuil))) +
  geom_density(position = "identity",
                 alpha = 0.5) +
  scale_fill_manual(values = c("purple", "yellow"))
```

### Calculation of IPTW weighting

The IPTW weighting is calculated from the propensity score and corresponds to the inverse of the propensity score.

If HLQ = 1 alors IPTW = 1/PS and if HLQ =0 then,IPTW = 1/(1-PS)

```{r}
#Calculation of frequency to calculate stabilized weights
proba_exposure <- sum(data$SEMSA_seuil == 1) / nrow(data)

#Calculation of IPTW : inverse of the propensity score according to HLQ_moy_bin ><1
data <- data %>%
  mutate(iptw = case_when(
  SEMSA_seuil == 1 ~ proba_exposure/data$propensity_score,
  SEMSA_seuil == 0 ~ (1-proba_exposure)/(1-data$propensity_score)
  ))

#Round the weight values to integers
head(data$iptw)
hist(data$iptw)
```

### Verification of the distribution of IPTW scores in the population

```{r}
#Verification of weight distribution 
summary(data$iptw)

#Visualisation
data %>%
  ggplot(aes(x = iptw, fill = as.factor(SEMSA_seuil))) +
  geom_density(position = "identity",
               alpha = 0.5) +
  scale_fill_manual(values = c("purple", "yellow"))
```

### Estimation of the estimand

Here, the estimand is the ATE due to the use of the IPTW. It answers the research question: what is the adjusted effect of the covariates (age, gender, BMI) on the probability of treatment adherence function of the HLQ score?

We use binary regression to estimate treatment adherence (y = final outcome) function of the SEMSA score, based on the previously calculated weights.

```{r}
##ogistic regression with new weights to estimate the effect of the HLQ score on binary variable adherence
fit <- glm(obs_J365_1M_PPC ~ SEMSA_seuil, 
           data = data,
           family = gaussian(),
           weights = iptw)
summary(fit)
```

Here the p-value is \< 0.05 which shows that the model is statistically significant for predicting treatment adherence from the binary SEMSA_seuil. Patients with SEMSA_seuil = 1 got exp(07882) = 2.2 more odds to be adherent to treatment.
