---
title: "Code PS IPTW ordinal variable"
author: "Jeanne LECLERE"
date: today 
format: html
editor: visual
warning: FALSE
message: FALSE
---

### Packages

```{r}
library("dplyr")
library("tidyverse")
library("ordinal")
library('ggplot2')
library('cobalt')
library('gtsummary')
library('brant')
```

## Data formatting and creation of an ordinal variable

```{r}
#Opening file
data <- read.csv("donnees_synthetiques.csv", header = TRUE, sep=",") 

#Search for values corresponding to percentiles to create categories
data |>
  tbl_summary(
    include = c(SEMSA_TOT),
    statistic = 
      all_continuous() ~ "Percentile : {p33} {p66} {p99} ")
p33 <- quantile(data$SEMSA_TOT, probs = 0.33, na.rm = TRUE)
p66 <- quantile(data$SEMSA_TOT, probs = 0.66, na.rm = TRUE)

#Creation of the HLQ categorical score variable: 2,3,4 becomes bad, average, good
data <- data %>%
  mutate(
    SEMSA_ord = case_when(
  SEMSA_TOT <= p33 ~  "1",
  SEMSA_TOT > p33 & SEMSA_TOT <= p66 ~ "2",
  SEMSA_TOT > p66  ~ "3")
  )

#Transform character in ordinal with 3 levels in the following order 1<2<3
data$SEMSA_ord <- factor(data$SEMSA_ord,ordered = TRUE)

#Verification of the distribution
table(data$SEMSA_ord)
```

### Calculation of the ordinal regression of the HLQ score function of covariates to calculate the propensity score

The score is calculated using ordinal logistic regression (the HLQ variable is transformed into an ordinal variable, resulting in scores of 2 -\> 1, 3 -\> 2, 4 -\> 3, etc.). It corresponds to the conditional probability (y) of exposure, weighted by the presence of covariates (x) (defined by DAG).

Here, the chosen covariates are age, gender, and BMI, as examples.

```{r}
#Calculating PS using ordinal logistic regression
ps_model <- MASS::polr(SEMSA_ord ~ AGE + GENDER + BODY.MASS.INDEX + APNEA.HYPOPNEA.INDEX,
                data = data, Hess = TRUE)
summary(ps_model)
```

#### Interpretation of regression results

Odds ratios are calculated comparing the probability that Y ≥ B to the probability that Y ≤ A.

Threshold between category 1 and 2 (1\|2): ; Threshold between category 2 and 3 (2\|3)

We observe that increasing age increases the OR: exp(0.021) = 1.024, meaning that the older the individual, the higher the probability of having a better HLQ score.

For gender, the OR = exp(-0.54) = 0.7, where 1 represents males, meaning men have a lower probability of having a low HLQ score.

For BMI, the OR exp(0.014) = 1.017, meaning the probability of having a better HLQ score increases with weight.

#### Slope test verification

For ordinal regression, it is necessary to verify the proportionality of the odds ratios by testing the slopes to check if the coefficients are equal between the k thresholds.

```{r}
#Verification of the proportionality of the odds ratios by applying a slope test
t <- brant::brant(ps_model)
```

Here all the tests are \> 0.05, so we do not reject H0 for any of the variables, which are all proportional. The transition between each threshold is constant (1-\>2-\>3).

### Calculating the propensity score using equations

We can therefore use our regression to calculate propensity scores (PS) from the regression equations, obtaining an equation of the form:

log(P(Y≤k)/P(Y\>k)) = θk - β1\*AGE + β2\*GENDER + β3\*BMI

where k represents two thresholds (Threshold 1 compares 1 vs. 2 and 3 / Threshold 2 compares 1 and 2 vs. 3). From this, we calculate the propensity score of each individual added to the database:

PS = e\^logit / (1+e\^(logit))

We calculate the probability of being in each category for each patient in order to estimate the entire distribution according to the chosen covariates.

```{r}
#Calculation of category probabilities: coefficients
beta <- coef(ps_model)

#Isolate the intercepts
intercept <- ps_model$zeta

#Isolate the coefficients of the covariates
betaX <- beta["AGE"] * data$AGE + beta["GENDER"] * data$GENDER +
        beta["BODY.MASS.INDEX"] * data$BODY.MASS.INDEX + beta["APNEA.HYPOPNEA.INDEX"] * data$APNEA.HYPOPNEA.INDEX

#Cumulative probabilities
p_mauvais_moyen <- 1/(1+exp(-(intercept["1|2"] - betaX)))

p_moyen_bon <- 1/(1+exp(-(intercept["2|3"] - betaX)))

#Probabilities per category: the sum of the scores = 1 for each individual
data$ps_mauvais <- p_mauvais_moyen
data$ps_moyen <- p_moyen_bon - p_mauvais_moyen
data$ps_bon <- 1 - p_moyen_bon 

summary(data$ps_mauvais)
summary(data$ps_moyen)
summary(data$ps_bon)

#Isolate PS for each calculated individual
data <- data %>%
  mutate(
    ps_ind = case_when(
  SEMSA_ord == "1" ~ data$ps_mauvais,
  SEMSA_ord == "2" ~ data$ps_moyen,
  SEMSA_ord == "3" ~ data$ps_bon)
  )
head(table(data$ps_ind))

#Visualization of PS according to the categories       
data %>%
  ggplot(aes(x = ps_ind, fill = SEMSA_ord)) +
  geom_density(alpha = 0.5) +  scale_x_log10() +
  scale_fill_manual(values= c("1"="purple","2"="yellow","3"="blue"))
```

### Calculating the propensity score using the predict function

```{r}
#Calculating the propensity score using predict = probability of being in each category for all individuals
table_ps <- data.frame(predict(ps_model, data, type = "probs")) 
head(table_ps)

#Verification of PS distribution
summary(table_ps)
table_ps <- table_ps %>% rename("1" = "X1", "2" = "X2", "3" = "X3")

#Add the PS to the database
data <- cbind(data, table_ps)

#Visualization of PS according to the categories 
#Pivot columns to visualize the data
data_graph <- data  %>% 
  select("1", "2", "3") %>%
  pivot_longer(cols = everything(), names_to = "Categories", values_to = "Score")
head(as.data.frame(table(data_graph)))

ggplot(data_graph, aes(x = Score, fill = Categories)) +
  geom_density(alpha = 0.5) +  scale_x_log10() +
  scale_fill_manual(values= c("1"="purple","2"="yellow","3"="blue"))
```

### Calculation of weighting IPTW

The IPTW score is calculated from the propensity score and is its inverse.

For an ordinal variable, the weight depends on the probability of the observed category; it corresponds to w_i = 1 / (P(T_i = t_i \| X_i) =\> 1/PS calculated previously.

```{r}
#Weight calculation = inverse of propensity score with individual weight. DO NOT ROUND.
data <- data %>%
  mutate(
    iptw = case_when(
  SEMSA_ord == "1" ~ 1/data$ps_ind,
  SEMSA_ord == "2" ~ 1/data$ps_ind,
  SEMSA_ord == "3" ~ 1/data$ps_ind)
  )
head(table(data$iptw))

#Visualization of the balance of covariances after weighting
graph <- bal.tab(SEMSA_ord ~ AGE + GENDER + BODY.MASS.INDEX + APNEA.HYPOPNEA.INDEX,
                 weights = data$iptw,
                 threshold = 0.1,
                 data = data,
                 un = TRUE)
print(graph)
love.plot(graph)
```

### Verification of the distribution of IPTW scores in the population

```{r}
#Verification of weight distribution
summary(data$iptw)

#Visualization 
data %>%
  ggplot(aes(x = iptw, fill = SEMSA_ord)) +
  geom_density(alpha = 0.5) +  scale_x_log10() +
  scale_fill_manual(values= c("1"="purple","2"="yellow","3"="blue"))
```

### Estimation of estimand

Here, the estimand is the ATE due to the use of the IPTW. It answers the research question: what is the adjusted effect of the covariates (age, gender, BMI) on the probability of treatment adherence as a function of the SEMSA score.

We use binomial regression to estimate treatment adherence (y = final outcome) as a function of the SEMSA score, based on the previously calculated weights.

```{r}
#Regression with the new weights using binary regression to determine if membership = yes/no
fit <- glm(obs_J365_1M_PPC ~ SEMSA_ord, #effect of the HLQ score on binary variable adherence
           data = data,
           family = gaussian(), #corresponding to Y type
           weights = iptw)

summary(fit)
```

Here we obtain two p-values, one for the linear effect "L" and one for the quadratic effect "Q" because the SEMSA_ord variable is ordered: 1 \< 2 \< 3.

The p-value is \< 0.05 for the linear effect, which shows that the model is statistically significant for a linear tendancy treatment adherence based on the ordinal SEMSA score.

The p-value is \> 0.05 for the quadratic effect, which shows that the model is not statistically significant for predicting treatment adherence based on the ordinal SEMSA score.
